\papertitle

### 2014-03-20 final-report/paper.tex
reworked some parts

\system's architecture (Figure~\ref{fig:arch}) consists of a back-end,
used as a plug-in interface for the distributed database system targeted, and a
front-end which implements the views described above. The front-end and
back-end of \system are highly decoupled so that the visualization can be used not only with
Myria but also with other systems such as Spark or Hadoop. In
Section~\ref{sec:back} we illustrate how we implemented log collections and
aggregation in Myria.

@@

The task of the back-end is to log when operators are called, when the call returns and when data is sent to another worker. Using this data and simple transformations, we can obtain the information necessary to create the visualizations to answer the questions in Section~\ref{sec:approach}.

There are two possible approaches that we explored to collect event logs that can be used to extract spans and other performance related data necessary for the visualization. First, events could be logged to files using a standard logging tool and then collect the files on one node using remote copy. Once the logs are on one machine, the logs can be parsed and the data can be transformed into the desired format.

We switched to a different approach as shown in Figure~\ref{fig:arch} where logs are written directly into the worker's database. When the front-end requests the data, the master executes a query that collects the data and streams the results directly to the requesting client. Logs are thus tuples in a relation and no longer separate in files and in a custom format. Despite being more reliable, faster and using the same logic that Myria already implements, we are able to write queries to aggregate, filter and transform the collected data.

For example, the data for how many workers are working on a certain fragment as described in Section~\ref{sec:fragments} is generated from the event log using a query to select only events for the root operators (the operators that have no parent in the same fragment) in each fragment and a custom map function that carries state. The state is an integer that is incremented when the operator is called on a worker and decremented when the operator returns.

@@

\subsection{Front-end}

The Myria web front-end server is written in Python and runs on Google App Engine\footnote{\url{https://developers.google.com/appengine/}}. \system's user-interface (UI) is embedded into Myria's web front-end. We build \system's UI using D3\cite{2011-d3} to deliver an interactive visualization experience to the user. D3 is a JavaScript framework for data visualization on the web. We used visualization techniques such as \emph{focus+context}\cite{furnas1986generalized} that possible to implement conveniently using D3.

\system's front-end gets data from the Myria server as JSON and CSV files. JSON objects are used to describe the physical queryplan. CSV files contain logs on query execution at the fragment and operator level as well as data exchange between workers at different stages of the query execution.

@@

\textbf{Graph Rendering:} We use D3 \cite{2011-d3} to render the \graph graph, which allows us to support various interactions and transitions. We use GraphViz \cite{Ellson01graphviz} in the backend to generate graph layout information. GraphViz allows us to generate graph layouts that are optimized to minimize area footprint. The layout information is then fed into a D3-based rendering engine which supports various interactions and animations techniques.

\textbf{Graph Navigation:} A query plan can be arbitrarily large. Thus we offer two mechanisms that facilitate exploration of the graph for the user: (1) expanding/reducing fragments and (2) zooming. The first mechanism can reduce the size of a graph by a constant factor by collapsing operators that compose a fragment into a single fragment node. The user can click once on a collapsed fragment to expand, and click once on an expanded fragment to collapse it. Expanding or reducing a node can cause large changes in the graph layout as GraphViz changes the graph layout to minimize overall area. This graph re-shuffling is illustrated in Figure~\ref{graph} where expanding Frag2 causes the layout to change. To address these layout changes, we implemented transitions to allow the user to track the fragments as those get reshuffled. The second mechanism that facilitates exploration of the graph for the user is zooming. This feature was implemented in D3 and allows the user to quickly jump from one part of the graph to the next, by zooming in and out of the graph view.

@@

The \fragment view comprises two charts: the utilization chart and the
operators chart. The utilization chart (Figure~\ref{fig:utilization_chart}) shows on
how many workers a fragment is running over time. The chart is used to
quickly reveal any patterns in the schedule, like tail latency, long periods of
idleness, etc. Execution skew is visible as humps.
We implemented a small brush (at the top of the figure) that
allows the user to zoom-in and further analyze the problematic area.

@@

\todo{use the word spans when referring to events with duration (used in dapper)}

@@

The \overall shows small multiples of the utilization chart as described in the previous section.
In addition to identifying execution skew, the user can compare the fragments and find
correlations between the execution of different fragments. This view helps the user navigate to the
right fragment and is not intended to offer further inside into the causes or performance problems.

@@

\todo{is this figure necessary or is the figure in the evaluation section enough?}

@@

We explore a simple join query we performed on one million tuples extracted from the Twitter user graph. This query joins Twitter followers and followees by user ID. The query can be written in SQL as:

@@

The \overall allows the user to quickly get an overview over the execution of all fragments
on all workers in the form of small multiples of the utilization charts as introduced in
Section~\ref{sec:fragment}. In Figure~\ref{fig:overview_skew} the user can see for example
that fragment~2 is very irregularly executed on different workers. If the graph is not a
single hill (as it is the case in Fragment 1, 3 and 4), workers are sleeping because they
either don't have enough data available to execute the operators or the output buffers
are full and data cannot be sent to the next fragment in the query graph.

The user has not identifies Fragment~2 to be a performance bottleneck and can use the other views
to further investigate its causes.

\begin{figure}[ht]
  \includegraphics[width=\columnwidth]{images/overview_skew}
  \caption{This \overall shows very irregular usage in fragment 2.}
  \label{fig:overview_skew}
\end{figure}

@@

\todo{Let's not call the fragments frag2 but write fragment 2}

The \network visualization at a glance shows how much data was sent between workers by one query fragment to the next. The two \network views in Figure~\ref{fig:network} were obtained by selecting the Frag0$\rightarrow$Frag2 and Frag1$\rightarrow$Frag2 edges respectively in the \graph view. Figure~\ref{fig:network} displays aggregate data communication between workers. In Figure~\ref{fig:no-skew} the amount of data sent from each worker to every other worker is balanced as opposed to Figure~\ref{fig:skew} where one can see that the volume of data sent to worker~1 and worker~5 outweighs data sent to the other workers in the system. This apparent data skew could be caused by partitioning issues.

@@

The \fragment visualization allows the user to dive down into query execution details at the worker level. The \fragment view in Figure~\ref{fig:exec_skew} was obtained by selecting Frag2, which is the fragment that performs the join, in the \graph view. The user can observe at a glance that certain workers produce much more data as a result of the join execution. Particularly, we saw in Case 2 Figure~\ref{fig:skew}, that tuples sent to Fragment 2 from Fragment 1 were heavily skewed towards being sent to worker~1 and worker~5. In Figure~\ref{fig:exec_skew}, the user can now see that as a result of the join, worker~5 ends up producing most tuples that have to be written back to disk. The color scheme used for each worker schedules in the \fragment view is color-coded to match the operators in Figure~\ref{fig:join}. Consequently, one can quickly observe in the \fragment view that worker~5 is busy sending join results from $t = 2.5s$ to $t = 5.0s$, which differs from the rest of the workers. Using similar reasoning, one can conclude that worker 5 presents a performance bottleneck in the execution of the query on Myria. In the short term, the user can modify the way the input data set gets partitioned. Alternatively she could change the number of workers allocated for the job. in the long term, developers should look into ways to improve data flow to avoid that workers have to wait for input data to process.

@@

All in all, \system allows the user to get much more insight into the execution of a query than what was available before, i.e. text-based performance logs, and runtimes. Being able to navigate performance data using a simple interface can indeed facilitate the task of a developer who is designing Myria's back end, or a Myria user who is writing complex queries to process large amounts of data. That said, \system can only help the user visualize query execution performance, but won't necessarily point the user to what she needs to do to improve performance of her query execution.

>>>>

\system's architecture (Figure~\ref{fig:arch}) is composed of: (1) a back-end
used as a plug-in interface for the targeted distributed database system, and (2) a
front-end that produces performance visualizations what we describe in Section~\ref{sec:front}. The front-end and
back-end of \system are highly decoupled so that the visualization can be used not only with
Myria but also with other systems such as Spark or Hadoop. In
Section~\ref{sec:back} we illustrate how we implemented log collections and
aggregation in Myria.

@@

The role of the back-end is produce event logs that record when operators are called, when the call returns and when data is sent to another worker. These logs are then used to create the visualizations in \system's web UI as described in Section~\ref{sec:front}.

There are two possible approaches that we explored to collect event logs. The first and simplest approach was to write the logs on files directly to disk. The files would then get hauled back on one node using remote copy. The logs would then be parsed, and the data transformed into the desired format.

We switched to a more scalable data collection approach approach as shown in Figure~\ref{fig:arch} where logs get written directly into the worker's database as tuples in a relation. When the front-end client requests data, the master executes a query that collects the data and streams the results directly to the requesting client. We found this data collection approach to be more scalable, reliable and faster. More particularly, we are able to write queries to aggregate, filter and transform the collected data.

For example, when querying how many workers are working on a certain fragment at a given time to produce the visualization described in Section~\ref{sec:fragments}, we used a query to select only events in the root operators (the operators that have no parent in the same fragment) for each fragment and a custom map function that carries state. The state is an integer that is incremented when the operator is called on a worker and decremented when the operator returns.

@@

\subsection{Front-end}
\label{sec:front}

The Myria web front-end server is written in Python and runs on Google App Engine\footnote{\url{https://developers.google.com/appengine/}}. \system's user-interface (UI) is embedded into Myria's web front-end. We build \system's UI using D3\cite{2011-d3} to deliver an interactive visualization experience to the user. D3 is a JavaScript framework for data visualization on the web. We used visualization techniques such as \emph{focus+context}\cite{furnas1986generalized} that were made possible using D3.

\system's front-end gets data from the Myria server using \system's back-end as JSON and CSV files. JSON objects are used to describe the physical queryplan. CSV files contain logs on query execution at the fragment and operator level as well as data exchange between workers at different stages of the query execution.

@@

\textbf{Graph Rendering:} We use D3 \cite{d3} to render the \graph graph, which allows us to support various interactions and transistions. We use GraphViz \cite{Ellson01graphviz} in the backend to generate graph layout information. GraphViz allows us to generate graph layouts that are optimized to minimize area footprint. The layout information is then fed into a D3-based rendering engine which supports various interactions and animations techniques.

\textbf{Graph Navigation:} A query plan can be arbitrarily large. Thus we offer two mechanisms that facilitate exploration of the graph for the user: (1) expanding/reducing fragments and (2) zooming. The first mechanism can reduce the size of a graph by a constant factor by collapsing operators that compose a fragment into a single fragment node. The user can click once on a collapsed fragment to expand, and click once on an expanded fragment to collapse it. Expanding or reducing a node can cause large changes in the graph layout as GraphViz changes the graph layout to minimize overall area. This graph reshuffeling is illustrated in Figure~\ref{graph} where expanding Frag2 causes the layout to change. To address these layout changes, we implemented transitions to allow the user to track the fragments as those get reshuffled. The second mechanism that facilitates exploration of the graph for the user is zooming. This feature was implemented in D3 and allows the user to quickly jump from one part of the graph to the next, by zooming in and out of the graph view.

@@

The \fragment view comprises two charts: the utilization chart and the
operators chart. The utilization chart (Figure~\ref{fig:utilization_chart}) shows on
how many workers a fragment is running over time. The chart is used to
quickly reveal any patterns in the schedule, like tail latency, long periods of
idleness, etc. We implemented a small brush (at the top of the figure) that
allows the user to zoom-in and further analyze the problematic area.

@@

% use the word spans when referring to events with duration (used in dapper)

@@

The \overall shows small multiples of the utilization chart as described in the previous section.
Execution skew can be spotted in this view. Furthermore, the user can compare the fragments and find
correlations between the execution of different fragments. This view helps the user navigate to the
right fragment and is not intended to offer further inside into the causes or performance problems.

@@

@@

We explore a simple join query we performed on one million tuples extracted from the Twitter user graph. This query joins Twitter followers and followees by user ID. The query can be written in SQL as expressed below.

@@

%dom

@@

The \network visualization at a glance shows how much data was sent between workers by one query fragment to the next. The two \network views in Figure~\ref{fig:no-skew} and Figure~\ref{fig:skew} were obtained by selecting the Frag0$\rightarrow$Frag2 and Frag1$\rightarrow$Frag2 edges respectively in the \graph view. Figure~\ref{fig:network} displays aggregate data communication between workers. In Figure~\ref{fig:no-skew} the amount of data sent from each worker to every other worker is balanced as opposed to Figure~\ref{fig:skew} where one can see that the volume of data sent to worker~1 and worker~5 outweights data sent to the other workers in the system. This apparent data skew could be caused by partitioning issues.

@@

The \fragment visualization allows the user to dive down into query execution details at the worker level. The \fragment view in Figure~\ref{fig:exec_skew} was obtained by selecting Frag2, which is the fragment that performs the join, in the \graph view. The user can observe at a glance that certain workers produce much more data as a result of the join execution. Particularly, we saw in Case 2 Figure~\ref{fig:skew}, that tuples sent to Fragment 2 from Fragment 1 were heavily skewed towards being sent to worker~1 and worker~5. In Figure~\ref{fig:exec_skew}, the user can now see that as a result of the join, worker~5 ends up producing most tuples that have to be written back to disk. The color scheme used for each worker schedules in the \fragment view is color-coded to match the operators in Figure~\ref{fig:join}. Consequently, one can quickly observe in the \fragment view that worker~5 is busy sending join results from t=2.5s to t=5.0s, which differes from the rest of the workers. Using similar reasoning, one can conclude that worker 5 presents a performance bottleneck in the execution of the query on Myria. Perhaps the user can modify the way the input data set gets partitioned. Alternatively she could change the number of workers allocated for the job.

@@

All in all, \system allows the user to get much more insight into the execution of a query than what was available before, i.e. text-based performance logs, and runtimes. Being able to navigate performance data using a simple interface can indeed facilitate the task of a developper who is designing Myria's back end, or a Myria user who is writing complex queries to process large amounts of data. That said, \system can only help the user visualize query execution performance, but won't necessarily point the user to what she needs to do to improve performance of her query execution.
